---
title: "Clinical Data Management"
---


```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Introduction

```{r, fig.align='center', echo = FALSE, fig.alt= "Clinical Data Management, Written and assembled by: Kate Isaac & Carrie Wright, Contributors: Christopher  Battiston, David Hanauer, Guilherme Del Fiol, Harry Hochheiser, Jennifer Kelleher, Abby Robbertz, & Meghan McGrady", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1ivDTcLjb2078O0GemkSeCgC1jmxk4fMsiFQaPaer9mQ/edit?slide=id.g34e03a999ff_0_165#slide=id.g34e03a999ff_0_165")
```


## Motivation


This course is intended for researchers (including postdocs and students) with limited to intermediate experience with informatics research. The conceptual material will also be useful for those in management roles who are collecting data and using informatics pipelines. 

This course is intended to provide an overview of the different kinds of data commonly used in clinical research, the questions that can typically be asked with such data, as well as guidance for how to manage this data.


```{r for_individuals_who, fig.align='center', echo = FALSE, fig.alt= "For individuals who: Have clinical data and don’t know what to do with it. Want a basic overview of what the clinical data represent and what questions can be asked.Need to know the basics of how to manage clinical data properly", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1ivDTcLjb2078O0GemkSeCgC1jmxk4fMsiFQaPaer9mQ/edit?slide=id.g116525eff64_0_96#slide=id.g116525eff64_0_96")
```


## Topics Covered

```{r topics_covered, fig.align='center', echo = FALSE, fig.alt= "Concepts discussed in Clinical Data course: What does your clinical data type represent? How do you handle the clinical data securely and ethically? What types of analysis questions can one ask when it comes to different kinds of clinical data? Find resources, tools and tutorials to help you process and work with your data", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1ivDTcLjb2078O0GemkSeCgC1jmxk4fMsiFQaPaer9mQ/edit?slide=id.g11db7c97851_0_143#slide=id.g11db7c97851_0_143")
```


## Curriculum 

The course will cover key underlying principles and concepts in ethical data handling. 

```{r learning_objectives, fig.align='center', echo = FALSE, fig.alt= "Overall Course Learning Objectives. This course will demonstrate how to: List common sources and types of clinical data, Explain the unique aspects related to clinical data, Understand uses and the questions that can be asked with the major types of clinical data,Recognize major tools and methods for managing clinical data ethically and securely",  out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1ivDTcLjb2078O0GemkSeCgC1jmxk4fMsiFQaPaer9mQ/edit?slide=id.gd422c5de97_0_10#slide=id.gd422c5de97_0_10")
```

