---
title: Clinical Data Uses
---

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Clinical Data Uses

## Learning Objectives

## General uses of clinical data

### Cancer Prevention and Care

The near universal adoption of electronic health record (EHR) systems in the US has created unprecedented opportunities to improve cancer prevention and care. As described in previous chapters, EHR systems store comprehensive longitudinal records of a patient's interactions with a healthcare system, including data about:

 - demographics
 - conditions
 - family history
 - clinical notes
 - laboratory test results
 - medications
 - procedures
 - imaging
 - genetic test reports

These data can be accessed not only by health professionals, but also by patients through patient portals. EHR data can also be used to enable data-driven interventions such as provider- and patient-facing clinical decision support (CDS) and population health management (PHM).

```{r, fig.align='center', echo = FALSE, fig.alt= "Electronic health records include many different kinds of records on individuals over time, including, clinical notes, family history information, lab result, images, and medication information. The image shows data on the same individual over a period of time. ", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1ivDTcLjb2078O0GemkSeCgC1jmxk4fMsiFQaPaer9mQ/edit#slide=id.g338f75828af_0_7" )
```


CDS has been defined as tools that "provide clinicians, staff, patients, or other individuals with knowledge and person-specific information, intelligently filtered or presented at appropriate times, to enhance health and health care" [@Osheroff_2007]. Examples of widely adopted CDS tools with demonstrated effectiveness for cancer prevention, diagnosis, and care include:

  * provider and patient reminders for cancer screening
  * decision aids and shared decision-making tools (e.g., for prostate cancer treatment, lung cancer screening)
  * AI-supported diagnostic imaging
  * chemotherapy decision support
  * at-home symptom care

While CDS tools generally provide decision support focused on a specific patient at a time, PHM are strategies that target specific patient populations [@Swarthout_Bishop_2017]. PHM efforts generally consist of:

  * population algorithms that are applied over EHR and other data sources to identify individuals who are eligible for a specific healthcare intervention (e.g., colorectal cancer screening, tobacco cessation, HPV vaccination)
  * patient engagement strategies (e.g., patient portals, text messaging, chatbots, patient navigators) that provide education and access to those interventions
  * analytic tools that assess the effectiveness of the PHM program.

Several PHM programs have demonstrated to be effective in increasing the uptake of cancer prevention. For example, the colorectal cancer screening program at Kaiser Permanente uses digital (i.e., text messaging, patient portal), mailed, and patient navigation approaches to increase colorectal cancer screening by mailing Fecal Immunohistochemical Test (FIT) kits to patients' homes [@Gupta2020]. Also the Cancer Moonshot BRIDGE trial used the GARDE platform [@Bradshaw2022] (ITCR-funded) to identify candidates for genetic testing of hereditary cancer syndromes based on EHR data; and for patient outreach, pre- and post-test education via automated chatbots [@Kaphingst2024].

https://pubmed.ncbi.nlm.nih.gov/31951474/



```{r, fig.align='center', echo = FALSE, fig.alt= "Clinical Decision Support (CDS) can help patients and clinicians make decisions about an individual’s care, while Population Health Management (PHM) can help identify individuals for interventions and engagement. The image shows a single person getting a colorectal screening reminder for CDS and a population being identified for possibly needing colorectal screening for PHM.", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1ivDTcLjb2078O0GemkSeCgC1jmxk4fMsiFQaPaer9mQ/edit#slide=id.g338f75828af_0_28" )
```

Clinical Decision Support (CDS) can help patients and clinicians make decisions about an individual’s care, while Population Health Management (PHM) can help identify individuals for interventions and engagement. The image shows a single person getting a colorectal screening reminder for CDS and a population being identified for possibly needing colorectal screening for PHM. 

While some CDS and PHM approaches have been successfully adopted widely, emerging technologies such as the use of generative AI approaches to analyze diagnostic imaging, large language models (LLMs) to extract information from narrative texts (e.g., clinical notes), LLM-based chatbots to communicate with patients, and digital health tools such as home-based sensors are creating unprecedented opportunities for next generation CDS and PHM. These approaches have the potential to enable significant breakthroughs through the implementation of patient-tailored cancer prevention and care at a population scale. Nevertheless, substantial research is needed to ensure effective, fair, and equitable implementation of these CDS and PHM interventions.

## Types of questions that can be asked with clinical data

### Risk Prediction

Risk prediction in clinical research involves using data to assess the likelihood of certain outcomes or events occurring in patients. This could include predicting the risk of developing a particular disease, experiencing a specific complication, or responding to a treatment.

Data used for risk prediction can come from various sources, including:

  * Clinical Data: This includes patient demographics, medical history, laboratory results, and imaging studies.
  * Genetic Data: Genetic information, such as DNA sequencing results, can provide valuable insights into an individual's susceptibility to certain diseases.
  * Environmental and Lifestyle Data: Factors such as diet, exercise habits, smoking status, and environmental exposures can influence disease risk and may be included in risk prediction models.
  * Biomarkers: Biological markers indicative of disease or physiological processes can be used as predictors in risk models [@Bodaghi_Fattahi_Ramazani_2023].

Once relevant data is collected, statistical and machine learning techniques can be applied to develop predictive models. These models aim to identify patterns and relationships within the data that are associated with the outcome of interest. Common techniques include logistic regression, decision trees, random forests, support vector machines, and neural networks.

After the model is trained on a dataset, it can be validated using independent datasets to assess its performance and generalizability. Once validated, the model can be used to predict risk in new patients based on their individual characteristics and data.

Clinical prediction rules are a subset of risk prediction models, specific to clinical research.

Examples of risk prediction models ...

Risk prediction models are important because ...

Overall, risk prediction in clinical research allows healthcare professionals to identify individuals at higher risk of certain outcomes, enabling targeted interventions, personalized treatments, and more efficient resource allocation.

### Cohort identification for research

Clinical data plays a crucial role in cohort identification for research purposes. Researchers typically use electronic health records (EHRs), medical databases, or registries to identify cohorts based on specific criteria such as age, gender, medical conditions, treatments, medications, and outcomes. Advanced data mining and natural language processing techniques can also be employed to extract relevant information from unstructured data sources like clinical notes. Once cohorts are identified, researchers can analyze the data to study disease progression, treatment effectiveness, and outcomes.

Cohort identification is important regardless of research study type, but to provide specific examples:

* Research Design: Identifying cohorts allows researchers to design studies with appropriate inclusion or exclusion criteria. By selecting specific groups of individuals with similar characteristics or exposures, researchers can investigate hypotheses effectively.
* Clinical Insights: Cohort studies enable researchers to observe the natural history of diseases, track outcomes over time, and assess the effectiveness of interventions or treatments. Understanding how different factors influence disease progression or treatment response can inform clinical decision-making and improve patient care.
* Epidemiological Studies: Cohort identification is crucial for epidemiological research to understand the incidence, prevalence, and risk factors associated with diseases. By following cohorts over time, researchers can identify trends, patterns, and associations that contribute to our understanding of disease causation and prevention.
* Precision Medicine: Identifying cohorts based on genetic profiles, biomarkers, or other specific characteristics allows researchers to tailor treatments and interventions to individual patients. This approach, known as precision medicine, aims to optimize therapeutic outcomes while minimizing adverse effects.
* Healthcare Policy and Planning: Cohort studies provide valuable data for informing healthcare policies, resource allocation, and public health strategies. By identifying high-risk populations or groups with specific healthcare needs, policymakers can develop targeted interventions to improve health outcomes and reduce disparities.

### Case report forms

We already have text for this

### Clinical trials/studies

We already have text for this

### Retrospective analysis

Retrospective analysis involves the examination of pre-existing clinical data to answer specific research questions, explore hypotheses, and identify patterns or trends. This method leverages historical data collected from medical records, administrative databases, registries, electronic health records (EHRs), or other sources of clinical information. The types of questions that can be asked through retrospective analysis span a wide range of clinical and epidemiological domains. The questions typically focus on understanding patient characteristics, disease epidemiology, treatment outcomes, risk factors, healthcare utilization, and more. 

Here, we explore the different categories of questions that can be addressed using retrospective data analysis:

1. **Descriptive Questions**

Descriptive questions aim to summarize and describe the characteristics of a patient population, disease, or healthcare process. They provide a foundational understanding of a dataset and are often the starting point for more complex analyses. 

- **Patient Demographics and Characteristics**: What are the demographic profiles (age, gender, ethnicity, socioeconomic status) of patients diagnosed with a specific condition? What are the common comorbidities and risk factors in this population?
- **Disease Prevalence and Incidence**: What is the prevalence or incidence of a specific disease or condition in a particular population or geographic area over a defined period?
- **Clinical Presentation**: What are the most common presenting symptoms, clinical signs, or laboratory findings associated with a specific disease or condition?

These questions help in understanding the baseline characteristics and epidemiology of diseases, which is crucial for planning further studies or developing public health interventions.

2. **Treatment and Intervention Questions**

Retrospective data can be used to examine the real-world effectiveness, safety, and patterns of treatment and interventions. These questions are critical for understanding how treatments are applied in clinical practice and their outcomes:

- **Treatment Effectiveness**: What is the effectiveness of a specific treatment or intervention in reducing symptoms, improving quality of life, or achieving clinical outcomes in a real-world setting?
- **Adherence and Persistence**: What are the rates of adherence and persistence with a prescribed treatment regimen over time? What factors are associated with higher or lower adherence rates?
- **Comparative Effectiveness**: How does the effectiveness of one treatment compare to another in a similar population? What are the relative benefits and risks associated with different treatment options?

By examining historical data, researchers can gain insights into how different treatments perform outside the controlled environment of clinical trials, thereby informing clinical decision-making and guidelines.

3. **Outcome and Prognostic Questions**

Understanding patient outcomes and prognostic factors is central to retrospective analyses. These questions focus on the end results of healthcare practices and patient management, including survival, complications, and quality of life: 

- **Survival and Mortality**: What are the survival rates and mortality rates associated with a particular disease, condition, or treatment? What factors are associated with increased or decreased survival rates?
- **Complication Rates**: What are the rates and types of complications associated with specific diseases, surgeries, or treatments? Are there identifiable risk factors for these complications?
- **Prognostic Factors**: What are the key prognostic factors (e.g., age, stage of disease, comorbidities) that influence the outcomes of patients with a particular condition?

These questions help identify factors that influence patient outcomes, guiding clinicians in predicting disease progression and tailoring individual treatment plans.

4. Risk Factor and Predictive Modeling Questions

Retrospective analyses are often used to identify potential risk factors for diseases and develop predictive models. These questions aim to uncover associations between variables and outcomes to help predict future events:

- **Risk Factors for Disease**: What are the risk factors associated with developing a particular disease or condition? Are there demographic, genetic, behavioral, or environmental factors that significantly increase the risk?
- **Prediction Models**: Can we develop a predictive model to estimate the likelihood of disease progression, relapse, or adverse outcomes based on historical patient data?
- **Multivariable Analysis**: How do multiple factors interact to influence the risk of an outcome? Are there synergistic or antagonistic effects between different risk factors?

These questions are critical for developing tools that help clinicians assess risk and make more informed decisions in preventive care and early intervention.

5. **Healthcare Utilization and Cost Questions**
Understanding how healthcare resources are utilized and the associated costs is important for optimizing healthcare delivery. Retrospective data can provide valuable insights into patterns of care and resource allocation for questions such as:

- **Healthcare Resource Utilization**: What are the patterns of healthcare utilization (e.g., hospitalizations, emergency room visits, outpatient visits) for patients with a specific condition? How does utilization vary by patient demographics, disease severity, or geographic location?
- **Cost of Care**: What are the direct and indirect costs associated with managing a specific disease or condition? How do different treatment options impact overall healthcare costs?
- **Cost-Effectiveness**: How does the cost-effectiveness of one treatment or intervention compare to another? What are the economic impacts of different clinical management strategies?

These questions can inform healthcare policy, budget planning, and the development of strategies to improve cost-effectiveness and access to care.

6. **Quality of Care and Patient Safety Questions**

Retrospective analysis can be used to evaluate the quality and safety of care provided to patients, identifying areas for improvement and potential patient safety concerns, helping to answer questions such as:

- **Quality Indicators**: What are the rates of adherence to clinical guidelines, protocols, or standards of care in managing a specific condition? How does adherence impact patient outcomes?
- **Adverse Events and Safety**: What is the incidence of adverse events, medication errors, or healthcare-associated infections in a particular patient population? What factors are associated with higher or lower rates of adverse events?
- **Patient Satisfaction and Experience**: What are the reported levels of patient satisfaction with healthcare services? How does satisfaction vary by demographic or clinical characteristics?

These questions help in assessing healthcare quality, identifying gaps in care, and developing interventions to enhance patient safety and satisfaction.

7. **Temporal and Trend Analysis Questions**

Retrospective studies often involve analyzing temporal patterns to understand how diseases, treatments, or outcomes change over time. These questions help in recognizing trends and patterns that can inform future research and policy decisions:

- **Temporal Trends in Disease Incidence and Prevalence**: How have the incidence and prevalence of a specific disease changed over the past decade? Are there seasonal or cyclical patterns in disease occurrence?
- **Trends in Treatment Practices**: How have treatment practices and prescribing patterns evolved over time? Are there emerging trends in the use of certain medications or interventions?
- **Trends in Outcomes and Survival**: How have patient outcomes, such as survival rates or complication rates, changed over time with advances in treatment and healthcare delivery?

These questions are essential for monitoring public health trends, evaluating the impact of interventions, and planning future healthcare strategies.

8. **Subgroup and Stratified Analysis Questions**

Subgroup analyses can help uncover differences in disease characteristics, treatment responses, and outcomes among different patient groups. This type of analysis is particularly valuable for personalized medicine and addressing questions related to health disparities:

- **Differences by Demographic Group**: How do treatment outcomes, risk factors, or disease characteristics differ between various demographic groups (e.g., age, gender, race/ethnicity)?
- **Subgroup-Specific Effectiveness**: Does a particular treatment work better in one subgroup of patients compared to another? What factors might explain these differences?
- **Health Disparities**: Are there disparities in healthcare access, treatment outcomes, or disease burden across different socioeconomic or minority groups?

Questions focusing on subgroups are crucial for understanding variations in care and outcomes, identifying at-risk populations, and developing targeted interventions.


Retrospective analysis of clinical data allows researchers to ask a wide range of questions that provide valuable insights into patient characteristics, disease epidemiology, treatment effectiveness, outcomes, risk factors, healthcare utilization, and more. By carefully formulating questions and analyzing historical data, researchers can uncover patterns, identify trends, and generate evidence that informs clinical practice, healthcare policy, and future research directions. This chapter outlines the various categories of questions that can be addressed through retrospective analysis, highlighting the potential of this approach to advance knowledge and improve patient care in diverse clinical settings.


## Research Design Considerations

Researchers need to intentionally use methods earlier in the research process than data analysis to manage data biases associated with clinical data, especially EHR.

One of the most important challenges in using EHR data in cancer research is that, as in many other fields, healthcare data are plagued with several types of biases that result from disparities in the delivery of healthcare. Overall, cancer research has historically relied on data from high resource academic medical centers, which disproportionately provide care to patients who are White, have high socioeconomic status, and live in urban areas. As a result, medical knowledge produced from these data have disproportionately benefited those patients. Different sources of bias are prevalent in EHR data. For example,

* *Information representativeness bias* occurs when certain groups are disproportionately less present in the EHR because they have no contact with the healthcare system.
* *Information presence bias*, on the other hand, occurs when certain groups may be represented in the EHR, but have disproportionately less comprehensive healthcare data due to issues such as overall lower access to and use of healthcare service, lack of a primary care provider, lack of access to specialty care, and lack of access to digital resources (e.g., patient portals, home sensors, telehealth) that can be used to provide healthcare data.
* *Treatment biases* happen when certain groups receive disproportionate access to more advanced treatments, which is often determined by social drivers such as insurance, distance, health literacy, and socioeconomic status.
* *Algorithm bias* further amplifies these previous sources of bias by leveraging biased EHR data to make predictions about diagnosis, treatment and prognosis that are used by clinicians to make potentially biased healthcare decisions, which are then documented in the EHR.

Recent advances in sophisticated and costly technology such as genetic testing, artificial intelligence, and digital health, which are disproportionately available in high resource healthcare systems further compound the problem. Therefore, cancer researchers increasingly need to use intentional methods to prevent, identify, and correct for biases in EHR data. For example, the National Institutes of Health Pragmatic Trials Collaboratory has made several recommendations to address EHR data biases in research [@BoydCCT2023; @BoydJAMIA2023]:

* Include data from low resource healthcare settings such as community health centers that provide care for patients who are racially diverse, have low socioeconomic status and live in rural areas.
* Engage with diverse communities in study design and conduct to ensure proper representation, data collection, analysis, and representation.
* Use data collection methods for self-reported data that rely on more accessible technology such as text messaging, using accessible and culturally adapted communication.
* Include subgroup analysis by different demographic groups according to variables such as socioeconomic status, race, ethnicity, sex, geographical location, and social determinants of health.

With such approaches, cancer researchers while aiming to avoid exacerbating health disparities, in addition help to reduce disparities.

## Conclusion
